\LARGE \bf Mean Estimation from Adaptive One-bit Measurements

### 2017-09-28 Writeups/onebit_adaptive_Allerton_V2.tex
changes to onebit_Allerton_V2.tex

\subsubsection*{Proof of Lem.~\ref{lem:bound_intervals}}
We use induction on $n \in \mathbb N$. For the base case $n=1$ we have 
\begin{equation} \label{eq:induction_base}
\frac{  \phi^2(x)} 
{\Phi(x) \left(1 - \Phi(x) \right) }. 
\end{equation}
Taking the logarithm of \eqref{eq:induction_base} and differentiating, we conclude that any point $x$ that maximizes \eqref{eq:induction_base} also satisfies
\[
x = \phi(x) \left( \Phi(x) -\frac{1}{2} \right).
\]
However, since $x > \phi(x) \left( \Phi(x) -\frac{1}{2} \right)$ for all $x > 0$, the only point that satisfies the last condition is $x=0$, in which \eqref{eq:induction_base} equals $2/\pi$. \par
Assume now that \eqref{eq:bound_intervals} holds for all integers up to some $n = N-1$ and consider the case $n = N$. The maximal value of \eqref{eq:bound_intervals} is attained for the same $(x_1,\ldots,x_N) \in \mathbb R^N$ that attains the maximal value of 
\begin{align*}
& g(x_1,\ldots, x_N) \triangleq 2 \log \left(  \sum_{k=1}^{N} (-1)^{k+1} \phi(x_k) \right) - \\
&  \log
\left( \sum_{k=1}^N (-1)^{k+1} \Phi(x_k) \right)
-\log \left(1 -  \sum_{k=1}^N (-1)^{k+1} \Phi(x_k) \right) \\
& = 2 \log \delta_N - \log \Delta_N - \log \left(1 - \Delta_N  \right),
\end{align*}
where we denoted $\delta_N \triangleq \sum_{k=1}^{N} (-1)^{k+1} \phi(x_k)$ and $\Delta_N =  \sum_{k=1}^N (-1)^{k+1} \Phi(x_k)$. The derivative of $g(x_1,\ldots,x_N)$ with respect to $x_k$ is given by
\[
\frac{\partial  g}{\partial x_k} = \frac{2 (-1)^{k+1} \phi'(x_k)}{\delta_N} -\frac{(-1)^{k+1} \phi(x_k)}{\Delta_N } + \frac{(-1)^{k+1} \phi(x_k)}{1-\Delta_N }.
\]
Using the fact that $\phi'(x) = -x \phi(x)$, we conclude that the gradient of $g$ vanishes only if
\[
x_k = \frac{\delta_N}{2} \left( \frac{1}{\Delta_N} - \frac{1}{1-\Delta_N} \right),\quad k=1,\ldots,N.
\]
In particular, the condition above implies $x_1 = \ldots = x_N$. If $N$ is odd then for $x_1=\ldots =x_N$ we have that the LHS of \eqref{eq:bound_intervals} equals
\[
\frac{\phi(x_1)^2}{ \Phi(x_1) (1 - \Phi(x_1))},
\]
which was shown to be smaller than $\pi/2$. If $N$ is even, then for any constant $c$ the limit of the LHS of \eqref{eq:bound_intervals} exits as $(x_1,\ldots,x_N)\rightarrow (c,\ldots,c)$ and equals zero. Therefore, the maximum of the LHS of \eqref{eq:bound_intervals} is not attained at this line. We now consider the possibility that the LHS of \eqref{eq:bound_intervals} is maximized at the borders, as one or more of the coordinates of $(x_1,\ldots,x_N)$ approaches plus or minus infinity. For simplicity we only consider the cases where $x_N$ goes to minus infinity or $x_1$ goes to plus infinity (the general case where the first $m$ coordinates goes to infinity or the last $m$ to minus infinity is obtained using similar arguments). Assume first $x_N \rightarrow -\infty$. Then the LHS of \eqref{eq:bound_intervals} equals
\begin{align*}
\frac{ \left(  \sum_{k=1}^{N-1} (-1)^{k+1}\phi(x_k) \right)^2} 
{\left( \sum_{k=1}^{N-1} (-1)^{k+1} \Phi(x_k) \right)\left(1- \sum_{k=1}^{N-1} (-1)^{k+1} \Phi(x_k)  \right) } ,
\end{align*}
which is smaller than $2/\pi$ by the induction hypothesis. Assume now that $x_1 \rightarrow \infty$. Then the LHS of
\eqref{eq:bound_intervals} equals
\begin{align*}
& \frac{ \left(  \sum_{k=2}^{N} (-1)^{k+1}\phi(x_k) \right)^2} 
{\left( 1 + \sum_{k=2}^{N} (-1)^{k+1} \Phi(x_k) \right)\left(1- 1 - \sum_{k=2}^{N} (-1)^{k+1} \Phi(x_k)  \right) }  \\
& = \frac{ \left(  -\sum_{m=1}^{N} (-1)^{m+1}\phi(x'_m) \right)^2} 
{\left( 1 - \sum_{m=1}^{N-1} (-1)^{m+1} \Phi(x'_{m}) \right)\left( \sum_{m=1}^{N-1} (-1)^{m+1} \Phi(x'_{m})  \right) },
\end{align*}
where $x'_{m} = x_{m+1}$. The last expression is also smaller than $2/\pi$ by the induction hypothesis. This proves Lem.~\ref{lem:bound_intervals}. \\

@@

\subsection*{Proof of Thm.~\ref{thm:sgd}}
The algorithm given in \eqref{eq:sgd_alg} and \eqref{eq:sgd_est} is a special case of a more general class of estimation procedures given in \cite{polyak1992acceleration}. In particular, Thm.~\ref{thm:sgd} is follows directly from the following simplified version of \cite[Thm. 4]{polyak1992acceleration}:
\begin{thm}{\cite[Thm. 4]{polyak1992acceleration}} \label{thm:polyak_juditsky}
Let 
\[
X_i = \theta + Z_i,\quad i=1,\ldots,n,
\]
where the $Z_i$s are i.i.d. from $\mathcal N(0,\sigma^2)$ and independent of each other. Define 
\begin{align*}
\theta_i & = \theta_{i-1} + \gamma_i \varphi(X_i - \theta_{i-1}), \\
\widehat{\theta}_n & = \frac{1}{n} \sum_{i=0}^{n-1} \theta_i,
\end{align*}
where the sequence $\left\{ \gamma_i \right\}_{i=1}^\infty$ satisfies conditions (i) and (ii) in Thm.~\ref{thm:sgd}, and $\left| \varphi(x) \right| \leq K_1(1+x)$ for some $K_1$. Define $\psi(x) = \mathbb E \varphi(x+Z_1)$, $\chi(x) = \mathbb E \varphi^2(x+Z_1)$ and assume that $\psi(0)=0$, $x\psi(x) >0$ for all $x\neq 0$, $\chi(x)$ is continuous at zero, and that $\psi(x)$ is differentiable at zero with $\psi'(0)>0$. Moreover, assume that there exists $K_2$ and $0<\lambda \leq 1$ such that
\[
\left| \psi(x) - \psi'(0)x \right|\leq K_2 |x|^{1+\gamma}.
\]
Then $\widehat{\theta}_n \rightarrow \theta$ almost surely and $ \sqrt{n}(\widehat{\theta}_n - \theta)$ converges in distribution to $\mathcal N(0,V)$, where
\[
V = \frac{ \chi(0)} {\psi'^2(0)}. 
\]
\end{thm}

Using the notation in Thm.~\ref{thm:polyak_juditsky}, we set $\varphi(x) = \sgn(x)$ and $Z_i = X_i - \theta$. We have: 
\[
\chi(x) = \mathbb E \varphi^2(x+Z_1),
\]
so $\chi(0) = 1$. In addition,
\begin{align*}
\psi(x) & = \mathbb E \sgn(x+ Z_1) = \int_{-\infty}^\infty \sgn(x+z) \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{\sigma^2}} dz \\
& = \int_{-x}^\infty \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{\sigma^2}} dz -\int_{-\infty}^{-x} \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{\sigma^2}} dz.
\end{align*}
This leads to 
\begin{align*}
\psi'(x) & = \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{x^2}{\sigma^2}} dz +\frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{\sigma^2}} dz,
\end{align*}
so $\psi'(0) = \frac{2}{\sqrt{2\pi}\sigma}$. It is now easy to verify that the rest of the conditions in Thm.~\ref{thm:polyak_juditsky} are fulfilled for any $\lambda > 0$. Since 
\[
\frac{\chi(0)}{\psi'^2(0)} = \frac{\pi \sigma^2}{2},
\]
Thm.~\ref{thm:sgd} follows from Thm.~\ref{thm:polyak_juditsky}. \QEDA

>>>>

\subsubsection*{Proof of Lem.~\ref{lem:bound_intervals}}
We use induction on $n \in \mathbb N$. For the base case $n=1$ we have 
\begin{equation} \label{eq:induction_base}
\frac{  \phi^2(x)} 
{\Phi(x) \left(1 - \Phi(x) \right) }. 
\end{equation}
Taking the logarithm of \eqref{eq:induction_base} and differentiating, we conclude that any point $x$ that maximizes \eqref{eq:induction_base} also satisfies
\[
x = \phi(x) \frac{\Phi(x) -\frac{1}{2}}{\Phi(x)(1-\Phi(x))} .
\]
However, since 
\[
x > \phi(x) \frac{\Phi(x) -\frac{1}{2}}{\Phi(x)(1-\Phi(x))}
\]
for all $x > 0$, the only point that satisfies the last condition is $x=0$, in which \eqref{eq:induction_base} equals $2/\pi$. \par
Assume now that \eqref{eq:bound_intervals} holds for all integers up to some $n = N-1$ and consider the case $n = N$. The maximal value of \eqref{eq:bound_intervals} is attained for the same $(x_1,\ldots,x_N) \in \mathbb R^N$ that attains the maximal value of 
\begin{align*}
& g(x_1,\ldots, x_N) \triangleq 2 \log \left(  \sum_{k=1}^{N} (-1)^{k+1} \phi(x_k) \right) - \\
&  \log
\left( \sum_{k=1}^N (-1)^{k+1} \Phi(x_k) \right)
-\log \left(1 -  \sum_{k=1}^N (-1)^{k+1} \Phi(x_k) \right) \\
& = 2 \log \delta_N - \log \Delta_N - \log \left(1 - \Delta_N  \right),
\end{align*}
where we denoted $\delta_N \triangleq \sum_{k=1}^{N} (-1)^{k+1} \phi(x_k)$ and $\Delta_N =  \sum_{k=1}^N (-1)^{k+1} \Phi(x_k)$. The derivative of $g(x_1,\ldots,x_N)$ with respect to $x_k$ is given by
\[
\frac{\partial  g}{\partial x_k} = \frac{2 (-1)^{k+1} \phi'(x_k)}{\delta_N} -\frac{(-1)^{k+1} \phi(x_k)}{\Delta_N } + \frac{(-1)^{k+1} \phi(x_k)}{1-\Delta_N }.
\]
Using the fact that $\phi'(x) = -x \phi(x)$, we conclude that the gradient of $g$ vanishes only if
\[
x_k = \frac{\delta_N}{2} \left( \frac{1}{\Delta_N} - \frac{1}{1-\Delta_N} \right),\quad k=1,\ldots,N.
\]
In particular, the condition above implies $x_1 = \ldots = x_N$. If $N$ is odd then for $x_1=\ldots =x_N$ we have that the LHS of \eqref{eq:bound_intervals} equals
\[
\frac{\phi(x_1)^2}{ \Phi(x_1) (1 - \Phi(x_1))},
\]
which was shown to be smaller than $\pi/2$. If $N$ is even, then for any constant $c$ the limit of the LHS of \eqref{eq:bound_intervals} exits as $(x_1,\ldots,x_N)\rightarrow (c,\ldots,c)$ and equals zero. Therefore, the maximum of the LHS of \eqref{eq:bound_intervals} is not attained at this line. We now consider the possibility that the LHS of \eqref{eq:bound_intervals} is maximized at the borders, as one or more of the coordinates of $(x_1,\ldots,x_N)$ approaches plus or minus infinity. For simplicity we only consider the cases where $x_N$ goes to minus infinity or $x_1$ goes to plus infinity (the general case where the first $m$ coordinates goes to infinity or the last $m$ to minus infinity is obtained using similar arguments). Assume first $x_N \rightarrow -\infty$. Then the LHS of \eqref{eq:bound_intervals} equals
\begin{align*}
\frac{ \left(  \sum_{k=1}^{N-1} (-1)^{k+1}\phi(x_k) \right)^2} 
{\left( \sum_{k=1}^{N-1} (-1)^{k+1} \Phi(x_k) \right)\left(1- \sum_{k=1}^{N-1} (-1)^{k+1} \Phi(x_k)  \right) } ,
\end{align*}
which is smaller than $2/\pi$ by the induction hypothesis. Assume now that $x_1 \rightarrow \infty$. Then the LHS of
\eqref{eq:bound_intervals} equals
\begin{align*}
& \frac{ \left(  \sum_{k=2}^{N} (-1)^{k+1}\phi(x_k) \right)^2} 
{\left( 1 + \sum_{k=2}^{N} (-1)^{k+1} \Phi(x_k) \right)\left(1- 1 - \sum_{k=2}^{N} (-1)^{k+1} \Phi(x_k)  \right) }  \\
& = \frac{ \left(  -\sum_{m=1}^{N} (-1)^{m+1}\phi(x'_m) \right)^2} 
{\left( 1 - \sum_{m=1}^{N-1} (-1)^{m+1} \Phi(x'_{m}) \right)\left( \sum_{m=1}^{N-1} (-1)^{m+1} \Phi(x'_{m})  \right) },
\end{align*}
where $x'_{m} = x_{m+1}$. The last expression is also smaller than $2/\pi$ by the induction hypothesis. This proves Lem.~\ref{lem:bound_intervals}. \\

@@

\subsection*{Proof of Thm.~\ref{thm:sgd}}
The algorithm given in \eqref{eq:sgd_alg} and \eqref{eq:sgd_est} is a special case of a more general class of estimation procedures given in \cite{polyak1992acceleration}. In particular, Thm.~\ref{thm:sgd} is follows directly from the following simplified version of \cite[Thm. 4]{polyak1992acceleration}:
\begin{thm}{\cite[Thm. 4]{polyak1992acceleration}} \label{thm:polyak_juditsky}
Let 
\[
X_i = \theta + Z_i,\quad i=1,\ldots,n,
\]
where the $Z_i$s are i.i.d. with zero mean. Define 
\begin{align*}
\theta_i & = \theta_{i-1} + \gamma_i \varphi(X_i - \theta_{i-1}), \\
\widehat{\theta}_n & = \frac{1}{n} \sum_{i=0}^{n-1} \theta_i,
\end{align*}
where the sequence $\left\{ \gamma_i \right\}_{i=1}^\infty$ satisfies conditions (i) and (ii) in Thm.~\ref{thm:sgd}, and $\left| \varphi(x) \right| \leq K_1(1+x)$ for some $K_1$. Define $\psi(x) = \mathbb E \varphi(x+Z_1)$, $\chi(x) = \mathbb E \varphi^2(x+Z_1)$ and assume that $\psi(0)=0$, $x\psi(x) >0$ for all $x\neq 0$, $\chi(x)$ is continuous at zero, and that $\psi(x)$ is differentiable at zero with $\psi'(0)>0$. Moreover, assume that there exists $K_2$ and $0<\lambda \leq 1$ such that
\[
\left| \psi(x) - \psi'(0)x \right|\leq K_2 |x|^{1+\gamma}.
\]
Then $\widehat{\theta}_n \rightarrow \theta$ almost surely and $ \sqrt{n}(\widehat{\theta}_n - \theta)$ converges in distribution to $\mathcal N(0,V)$, where
\[
V = \frac{ \chi(0)} {\psi'^2(0)}. 
\]
\end{thm}

Using the notation in Thm.~\ref{thm:polyak_juditsky}, we set $\varphi(x) = \sgn(x)$ and $Z_i = X_i - \theta$. We have: 
\[
\chi(x) = \mathbb E \varphi^2(x+Z_1),
\]
so $\chi(0) = 1$. In addition,
\begin{align*}
\psi(x) & = \mathbb E \sgn(x+ Z_1) = \int_{-\infty}^\infty \sgn(x+z) \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz \\
& = \int_{-x}^\infty \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz -\int_{-\infty}^{-x} \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz.
\end{align*}
This leads to 
\begin{align*}
\psi'(x) & = \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{x^2}{2\sigma^2}} dz +\frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz,
\end{align*}
so $\psi'(0) = \frac{2}{\sqrt{2\pi}\sigma}$. It is now easy to verify that the rest of the conditions in Thm.~\ref{thm:polyak_juditsky} are fulfilled for any $\lambda > 0$. Since 
\[
\frac{\chi(0)}{\psi'^2(0)} = \frac{\pi \sigma^2}{2},
\]
Thm.~\ref{thm:sgd} follows from Thm.~\ref{thm:polyak_juditsky}. \QEDA
